---
title: "PCA"
author: "Huub Hoefsloot"
date: "29 January 2019"
output: html_document
---

# Setup

## Warm up
Let's multiply two vectors
```{r}
t1 <- c(1,2,4,8,8)
p1 <- c(2,2,4)
t1 %*% t(p1)

```

## Library
```{r}
install.packages("tiff")
library("tiff")
```

## TIFF file
You should first download the `cameraman.tif' file on your local machine. 
```{r}
# put the tif file in your working directory
cam <- readTIFF('cameraman.tif', native = FALSE, all = FALSE, convert = FALSE, info = FALSE, indexed = FALSE, as.is = FALSE)
N <- 5
USV <- svd(cam)
T1 <- USV$u %*% diag(USV$d)
P1 <- USV$v
cam2 <- USV$u[,1:N] %*% diag(USV$d[1:N]) %*% t(USV$v[,1:N])

# a tif file should have value between 0 and 1 so I need to do a correction
for (i in 1:256){
     for (j in 1:256){
         if (cam2[i,j]<0) {cam2[i,j]=0}
         if (cam2[i,j]>1) {cam2[i,j]=1}
     }
 }
writeTIFF(cam2,"Z:/BDA/cam2.tif") # choose a proper destination on your machine
```

# PCA on random data

Let's make some random data and see how PCA results look if there is no information in the data.
```{r}
xvec <- rnorm(12*86)
Xrand=matrix(xvec,12,86)

m1= colMeans(Xrand) 				    #calculates column means
X1=sweep(Xrand,2,m1,FUN="-")		#subtract the column means
ssqtotal <- sum(X1*X1)			    # calculates the sum of squares
USV <- svd(X1)					        # see the help for this function
T <- USV$u %*% diag(USV$d)			# calculates the scores
P <- USV$v					            # calculates the loadings
npc <- 10						            # number of components
ssqtot <- 0 * (1:npc)
for (i in 1:npc){
  Xest  <- T1[,1:i] %*% t(P1[,1:i])
  ssqtot[i] <- 100 * (1-sum((X1-Xest)*(X1-Xest))/ssqtotal)
}

ssq=c(ssqtot[1],ssqtot[2:10]-ssqtot[1:9])
data.frame(ssq=ssq,ssqtot=ssqtot)
```

Plot the first score against the second score.
```{r}
plot(T[1:12,1],T[1:12,2], pch=3, col="blue", xlab= "T1",ylab ="T2", main = " scores")
points(T[7:12,1],T[7:12,2], pch=8, col="red")
```

Put an artifical effect into the random data.  
Do the PCA again.
```{r}
# add some artifial effect
Xrand[1:6,1:10] <- Xrand[1:6,1:10] + 5

# do the pca again
m1= colMeans(Xrand) 				#calculates column means
X1=sweep(Xrand,2,m1,FUN="-")		#subtract the column means
ssqtotal <- sum(X1*X1)			# calculates the sum of squares
USV <- svd(X1)					# see the help for this function
T1 <- USV$u %*% diag(USV$d)			# calculates the scores
P1 <- USV$v					# calculates the loadings
npc <- 10						# number of components
ssqtot <- 0 * (1:npc)
for (i in 1:npc){
  Xest  <- T1[,1:i] %*% t(P1[,1:i])
  ssqtot[i] <- 100 * (1-sum((X1-Xest)*(X1-Xest))/ssqtotal)
}
ssq=c(ssqtot[1],ssqtot[2:10]-ssqtot[1:9])
data.frame(ssq=ssq,ssqtot=ssqtot)
```

Plot the first score against the second score once again.
```{r}
plot(T[1:12,1],T[1:12,2], pch=3, col="blue", xlab= "T1",ylab ="T2", main = " scores")
points(T[7:12,1],T[7:12,2], pch=8, col="red")
```

# PCA on real data

## Load dataset
Load the dataset. These dataset gives metabolites concentrations for 7control plants and 7 plants placed in the dark.  
Make sure you indicate the correct file path on your system. 
```{r}
data <- read.csv(file='Caldana.csv') # indicate the correct path on your system
X <- as.matrix(data)
```

## PCA computation
Perform a Principal Component Analysis of the untransformed and log transformed data. In both cases center the data (for the case of the transformed data, first take the logarithm of the data, then mean center the data).
```{r}
m1= colMeans(X) 					   # calculates column means
X1=sweep(X,2,m1,FUN="-")		 # subtract the column means
ssqtotal <- sum(X1*X1)			 # calculates the sum of squares
USV <- svd(X1)					     # see the help for this function
T1 <- USV$u %*% diag(USV$d)	 # calculates the scores
P1 <- USV$v					         # calculates the loadings
npc <- 10						         # number of components
ssqtot <- 0 * (1:npc)
for (i in 1:npc){
  Xest  <- T1[,1:i] %*% t(P1[,1:i])
  ssqtot[i] <- 100 * (1-sum((X1-Xest)*(X1-Xest))/ssqtotal)
}
ssq=c(ssqtot[1],ssqtot[2:10]-ssqtot[1:9])
data.frame(ssq=ssq,ssqtot=ssqtot)
```
T1 are the scores, P1 the loadings, ssq shows information on the variance explained. The build in R function svd calculates a singular value decomposition 
Note the dataframe shows: PCnumber, Variance explained per component and cumulative variation explained.

## Focus on scores
To closely look at the difference between the not transformed data and the log transformed data we will take a look at the scores plot and at the residuals histogram plot of both sets.  
```{r}
plot(T1[1:12,1],T1[1:12,2], pch=3, col="blue", xlab= "T1",ylab ="T2", main = " scores")
points(T1[7:12,1],T1[7:12,2], pch=8, col="red")
```

## Focus on loadings
Make a plot of the loadings
```{r}
plot(P2[,1],P2[,2])
text(P2[,1],P2[,2],1:63)
```

To connect the names with the numbers you can use `names(data)`.  

Look at a number of metabolites and try to understand the relationship with the data. Look at Table 2 in the Eisner paper at the muscle loss column.   
Can you understand the relationship between the PCA model and the variation in the data from this Table?  


```{r}

```

